 

### 2021-07-15 vertical-thermal-review.Rmd
combined the two Tleaf versions
please let me know if there are parts to cut over here

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections), with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
While $T_{leaf}$ is most commonly within a few degrees of $T_{air}$, it is rarely exactly equal [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input through radiation versus heat lost to the environment [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].**
The dominant component of a leaf's energy balance is net radiation ($R_n$, including shortwave and longwave), and high radiation loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig. `r fig_leaf_T`a).
The dominant avenue of heat loss to the environment is latent heat flux ($\lambda E$), wherein the loss of energy required convert water from liquid to vapour state cools the leaf. 
Evaporation ($E$) is controlled primarily by stomatal and boundary layer conductances, $g_s$ and $g_a$, and by the water vapour concentration difference between the leaf surface and the bulk air.
Therefore, $T_{leaf}-T_{air}$ decreases with $g_a$, which in turn increases with wind speed [Fig. `r fig_leaf_T`b,@daudetWindSpeedLeaf1999a] and decreases with leaf size (Fig. `r fig_leaf_T`d);
decreases with $g_s$ (Fig. `r fig_leaf_T`e);
and decreases with with vapor pressure deficit, or increases with RH (Fig. `r fig_leaf_T`c).

@@

**version 1 (for rest of biophyscial drivers subsection-- coauthors, please let us know which is better):**

Solar and longwave radiation can dramatically elevate $T_{leaf}$ well above $T_{air}$ (Fig. `r fig_leaf_T`a) because the heat absorbed through high irradiance can be greater than heat lost through sensible heat (via $g_a$; *i.e.*, boundary layer conductance) and $\lambda E$ (via $g_s$ *ie.* stomatal conductance and $E$ ie evaporation)[@bonan_ecological_2016].

$g_a$ and $g_s$ are dominant factors in $T_{leaf}$ flux. $Ga$ increases with wind speed which also in turn increases $gs$, thereby reduces $T_{leaf}$-$T_{air}$ (Fig. `r fig_leaf_T`b). Therefore, high windspeeds in upper canopies such as in NEON sites with windspeeds ~4 ms (Fig. `r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ that are exposed to high radiation loads, upto ~ 4$^\circ$C (Fig. `r fig_leaf_T`b)[@bonan_ecological_2016; @PhotosyntheticAcclimationSimultaneous2004]. Low wind speeds typically decouple leaves from $T_{air}$, more so when solar radiation is high and $gs$ is low. Additionally, $T_{leaf}-T_{air}$ increases with humidity (Fig. `r fig_leaf_T`c) because reduced $ga$ stifles $E$ and $gs$ that dissapate heat [@songTreeSurfaceTemperature2020].

Latent heat loss through maximum $g_s$, cools the leaf and decreases $T_{leaf}$-$T_{air}$, allowing coupling between leaf and air. However in conditions of low soil moisture during drought (Fig. `r fig_leaf_T`), greater temperature than $gs$ thermal optimum, and air dryness causing an increase in VPD, $gs$ becomes limited, thereby increasing $T_{leaf}$-$T_{air}$ [@bonan_ecological_2016]. Limitation to $gs$ can also exaggerate effects of biophysical factors that contribute to energy inputs such as, solar radiation, humidity and $T_{air}$ (Fig. `r fig_leaf_T`)[@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @dongBiophysicalHomoeostasisLeaf2017].

Leaf size and physiology [see next section] influence $g_a$, $g_s$, leaf thermal capacitance (capacity to hold heat), and therefore also $T_{leaf}$-$T_{air}$. Smaller leaves have thinner boundary layer and higher $g_a$, that enables coupling with $T_{air}$ compared to large leaves that can be more decoupled from $T_{air}$ (Fig. `r fig_leaf_T` a). Therefore, smaller leaves with adequate water, will be most effective at cooling during high temperatures than larger leaves [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007;@fauset_differences_2018]. Additionally, under low raditation, large leaves can be cooler than small leaves due to greater $g_s$ [Fig. `r fig_leaf_T` d, @leighThickLeavesAvoid2012;@bonan_ecological_2016]. Sun leaves, such as in the upper canopies, have greater $gs$, due to greater stomatal density (Table 1) and lower thermal capacitance due to lower water potential than shade leaves. However, exposure to high irradiance in upper canopies which is a dominant source of energy input increases $T_{leaf}$-$T_{leaf}$ more than shade leaves that are shielded from biophysical conditions [@millerOnlySunlitLeaves2021]. However, because shade leaves have higher thermal capacitance and comparitevly lower $gs$, when exposed to an increase in radiation or $T_{air}$, $T_{leaf}$-$T_{air}$ can easily become more elevated than sun leaves (Fig. `r fig_leaf_T`a)[@schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].

**Within forests, vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
$T_{leaf}$ generally tracks $T_{air}$, where vertical gradient of $T_{air}$ depends on canopy structure and $LAI$ [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021]. Closed canopies, and high $LAI$ canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in sub canopies and the understory microclimate. The intensity of incoming radiation and $VPD$ in the upper canopy can limit $gs$, therefore, we would expect a stronger $T_{leaf}$ gradient in closed canopies where upper canopy leaves can be much warmer than below canopies layers [@millerOnlySunlitLeaves2021]. However, higher windspeeds and lower RH (Fig. `r fig_NEON_vertical`), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (reviewed below), give them lower sensitivity to intercepted radiation (Fig. `r fig_leaf_T`a) and higher rates of heat loss (Fig. `r fig_leaf_T`b-e) [@millerOnlySunlitLeaves2021].
On the other hand, open canopies, or lower canopy $LAI$ allow more turbulent air mixing and sunlight into below canopies, that can either neutralize $T_{leaf}$ gradient or in the case of 'canopy greenhouse effect' or even sunflecks, can elevate below canopy $T_{leaf}$-$T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b].

**version 2 (for rest of biophyscial drivers subsection-- coauthors, please let us know which is better):**

@@

**These basic biophysical principles shape the propensity for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.**
High radiation at the top of the vertical profile (Fig. `r fig_NEON_vertical`) implies that canopy leaves have much greater propensity for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig. `r fig_leaf_T`).
However, higher windspeeds and lower RH (Fig. `r fig_NEON_vertical`), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (reviewed below), give them lower sensitivity to intercepted radiation (Fig. `r fig_leaf_T`a) and higher rates of heat loss (Fig. `r fig_leaf_T`b-e).

**Within closed-canopy forests, vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
Under hot conditions, higher wind speeds and $g_a$ enable cooling, but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf} - T_{air}$ [Fig. `r fig_leaf_T`b, @martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]. 
Thus, understory leaves, which typically experience lower windspeeds (Fig. `r fig_NEON_vertical`), would retain more heat.
The higher-RH conditions of the understory (Fig. `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig. `r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].
Thus, understory leaves can remain cooler under lower radiation load, but their environment is not conducive to shedding excess heat, particularly under humid conditions, and their tendency towards lower leaf $g_s$ and thermal capacitance can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020]. 
In contrast, under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated more in the canopy than the understory (Fig.`r fig_leaf_T`).
Because sun leaves are exposed to higher irradiance and $VPD$, their $g_s$-limitation is steeper that that of shade leaves, resulting in greater increases in $T_{leaf}$ when $g_s$ is reduced [Fig. `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].

@@

Further, we would expect expect $Q_{10}$ to also vary with environmental conditions based on canopy-position effect gradient in respiration rates for Quercus rubra trees being much greater in mesic than in xeric sites [@xuSeasonalVariationTemperature2006] but further research is required to discern $Q_{10}$.

>>>>

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections) with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
While $T_{leaf}$ is most commonly within a few degrees of $T_{air}$, it is rarely exactly equal [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input from radiation minus heat lost to the environment [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].**
The dominant component of a leaf's energy balance is net radiation ($R_n$, including shortwave and longwave). High radiation loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig. `r fig_leaf_T`a).
The dominant avenue of heat loss to the environment is latent heat flux ($\lambda E$), wherein the loss of energy required to convert water from a liquid to a vapour state during evaporation and transpiration cools the leaf. 
Evapotranspiration ($E$) is controlled primarily by stomatal and boundary layer conductances, $g_s$ and $g_a$, and by the water vapour concentration difference between the leaf surface and the bulk air.
Therefore, $T_{leaf}-T_{air}$ decreases with $g_a$, which in turn increases with wind speed [Fig. `r fig_leaf_T`b,@daudetWindSpeedLeaf1999a] and decreases with leaf size (Fig. `r fig_leaf_T`d);
decreases with $g_s$ (Fig. `r fig_leaf_T`e);
and decreases with vapor pressure deficit, or increases with RH (Fig. `r fig_leaf_T`c).

@@

@@

**These basic biophysical principles shape the propensity for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.**
High radiation at the top of the vertical profile (Fig. `r fig_NEON_vertical`) implies that canopy leaves have much greater propensity for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig. `r fig_leaf_T`).
However, higher windspeeds such as in NEON sites with windspeeds ~4 ms (Fig. `r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ that are exposed to high radiation loads, upto ~ 4$^\circ$C (Fig. `r fig_leaf_T`b)[@bonan_ecological_2016; @PhotosyntheticAcclimationSimultaneous2004]. Collectively, windspeed, lower RH (Fig. `r fig_NEON_vertical`), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (reviewed below), give them lower sensitivity to intercepted radiation (Fig. `r fig_leaf_T`a) and higher rates of heat loss (Fig. `r fig_leaf_T`b-e).

**Vertical gradients**

**Within closed-canopy forests, vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
Under hot conditions, higher wind speeds and $g_a$ enable cooling, but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf} - T_{air}$ [Fig. `r fig_leaf_T`b, @martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]. 
Thus, understory leaves, which typically experience lower windspeeds (Fig. `r fig_NEON_vertical`), would retain more heat.
The higher-RH conditions of the understory (Fig. `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig. `r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].
Thus, understory leaves or shade leaves in a canopy can remain cooler under lower radiation load, but their environment is not conducive to shedding excess heat, particularly under humid conditions, and their tendency towards lower leaf $g_s$ and thermal capacitance (capacity to hold heat) can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002]. 
In contrast, under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated more in the canopy than the understory (Fig.`r fig_leaf_T`).
Because sun leaves are exposed to higher irradiance and $VPD$, their $g_s$-limitation is steeper that that of shade leaves, resulting in greater increases in $T_{leaf}$ when $g_s$ is reduced [Fig. `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].

**Vertical $T_{leaf}$ gradients can also vary with canopy structure (Fig. `r fig_leaf_T`).**
$T_{leaf}$ generally tracks $T_{air}$, where vertical gradients of $T_{air}$ depend on canopy structure and $LAI$ [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021]. Closed canopies, and high $LAI$ canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate. The intensity of incoming radiation and $VPD$ in the upper canopy can limit $g_s$. Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than below canopies layers [@millerOnlySunlitLeaves2021]. However, higher windspeeds and lower $RH$ (Fig. 2), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (reviewed below), give them lower leaf sensitivity to intercepted radiation (Fig. 3a) and higher increase rates of heat loss (Fig. 3b-e). In contrast, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into the understory, that can either neutralize a $T_{leaf}$ gradient or elevate below canopy $T_{leaf}-T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves, as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b].

@@

Further, we would expect expect $Q_{10}$ to also vary with environmental conditions based on the evidence of canopy-position effect gradient in respiration rates for Quercus rubra trees being much greater in mesic than in xeric sites [@xuSeasonalVariationTemperature2006].
